A Type System for Structured Aliasing in Parallel Programs

### 2012-10-29 doc/types-pldi2013/intro.tex
some small edits

To combat this problem, researchers have previously explored programming
systems that are capable of reasoning about the structure of data in
parallel programs.  Traditionally, these attempts have taken two approaches.   
The first approach relies on fully static analyses that
have no runtime overhead, but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap recursively into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related} for more discussion of related work).  The
second approach relies on a dynamic analysis of program data that doesn't
limit expressivity, but incurs runtime overhead.  Examples of this approach 
include transactional memory and thread-level speculation.

These two different approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach outlaws aliasing
by requiring a single static partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach enables arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

@@

Using our proofs of the soundess of the privilegs and coherence systems, we then 
prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendant of $t_1$ in the task hierarchy is non-interfering
with any descendant of $t_2$.  This theorem is the basis for correct, distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the large communication latencies in the target
class of machines.

>>>>

To combat this problem, researchers have previously explored programming
systems that are capable of reasoning about the structure of data in
parallel programs.  Traditionally, these attempts have taken one of two approaches.   
The first approach relies on static analyses that
have no runtime overhead, but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap recursively into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related} for more discussion of related work).  The
second approach relies on a dynamic analysis of program data that doesn't
limit expressivity, but incurs runtime overhead.  Examples of this approach 
include transactional memory and thread-level speculation.

These two different approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
by requiring a single static partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach enables arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

@@

Using our proofs of the soundess of the privilegs and coherence systems, we then 
prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then any descendant of $t_1$ in the task hierarchy is non-interfering
with any descendant of $t_2$.  This theorem is the basis for correct, distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the large communication latencies in the target
class of machines.

### 2012-10-29 doc/types-pldi2013/legion-types.tex
some small edits

\begin{abstract}
Aliasing introduces a significant impediment to reasoning about the
structure and usage of data in parallel programs.  Traditional approaches
have either disallowed aliasing to enable static analyses, or have permitted
aliasing at the cost of dynamic runtime analyses.  In this paper we introduce a
type system for the Legion programming model that uses a hybrid static and
dynamic approach to manage aliasing with minimal runtime overhead.  We prove 
that our type system can soundly reason about the privilege and coherence 
properties in Legion programs. Furthermore, we prove that type soundness 
enables a hierarchical, distributed scheduling algorithm for Legion programs 
even in the presence of aliasing. We show that an implementation of a Legion 
type checker improves the performance of Legion programs up to 71\% by 
eliding memory checks that the type system proves safe.  We also demonstrate 
that an implementation of the Legion runtime based on the hierarchical 
scheduling algorithm scales well for three real-world applications, achieving 
up to 62.5X speedup on 96 GPUs on the Keeneland supercomputer.
%Legion is a programming model that introduces
%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
%with complex memory hierarchies.  Logical regions are first-class 
%values in Legion, and programmers may {\em partition} regions in multiple different
%ways into subregions.  We present
%a type system with privileges and coherence properties for reasoning about logical regions.  
%We prove the soundness of the type system and use it to show
%that all memory accesses in Legion are safe.  Furthermore, we prove that
%type soundness enables a scalable, hierarchical
%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
%for Legion programs improves performance up to 71\% by eliding
%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
%runtime based on the hierarchical scheduling algorithm scales well for
%three real-world applications.

>>>>

\begin{abstract}
Aliasing introduces a significant impediment to reasoning about the
structure and usage of data in parallel programs.  Traditional approaches
have either disallowed aliasing to enable static analyses, or have permitted
aliasing at the cost of dynamic runtime analyses.  In this paper we introduce a
type system for the Legion programming model that uses a hybrid static and
dynamic approach to manage aliasing with minimal runtime overhead.  We prove 
that our type system can soundly reason about the privilege and coherence 
properties in Legion programs. Furthermore, we prove that type soundness 
enables a hierarchical, distributed scheduling algorithm for Legion programs 
even in the presence of aliasing. We show that an implementation of a Legion 
type checker improves the performance of Legion programs up to 71\% by 
eliding memory checks that the type system proves safe.  We also demonstrate 
that an implementation of the Legion runtime based on the hierarchical 
scheduling algorithm scales well for three real-world applications running
on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 
on the Keeneland supercomputer.
%Legion is a programming model that introduces
%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
%with complex memory hierarchies.  Logical regions are first-class 
%values in Legion, and programmers may {\em partition} regions in multiple different
%ways into subregions.  We present
%a type system with privileges and coherence properties for reasoning about logical regions.  
%We prove the soundness of the type system and use it to show
%that all memory accesses in Legion are safe.  Furthermore, we prove that
%type soundness enables a scalable, hierarchical
%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
%for Legion programs improves performance up to 71\% by eliding
%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
%runtime based on the hierarchical scheduling algorithm scales well for
%three real-world applications.
